Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-03-03 manuscript/climate_sensitivity_draft.Rmd
finished integrating comments in ...wjmEGAnp_AES

**Authors:** 
Ian R. McGregor^1,2^,
Ryan Helcoski^1^,
Norbert Kunert^1,3^,
Alan J. Tepley^1,4^,
Erika B. Gonzalez-Akre^1^,
Valentine Herrmann^1^,
Joseph Zailaa^1,5^,
Atticus Stovall^1,6,7^,
Norman A. Bourg^1^,
William J. McShea^1^,
Neil Pederson^8^,
Lawren Sack^9,10^,
Kristina J. Anderson-Teixeira^1,3^*

@@

Forests globally play a critical role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change [@friedlingstein_climatecarbon_2006]. An important aspect of this uncertainty lies with physiological responses of trees to drought [@kennedy_implementing_2019]. In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014]; [Dai et al. 2018](https://link.springer.com/article/10.1007/s40641-018-0101-6)), often despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015]. Droughts, intensified by climate change, have been affecting forests worldwide [@allen_global_2010], and are expected to continue as one of the most important drivers of forest change in the future [@allen_underestimation_2015]. Understanding and modeling forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought resistance, and the extent to which their influence is consistent across droughts. However, it has proved difficult to resolve factors affecting tree growth during drought with available forest census data, which only rarely captures extreme drought, and with tree-ring records, which rarely represent many species and size classes at a single site and rarely sampled in association with detailed ecological data. The aim of this study was to pair tree-ring and forest plot data to test how tree size and traits shaped growth responses to historical droughts.

@@

Debates have also arisen regarding the traits influencing tree growth responses to drought. Commonly-measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011] and other forest biomes around the world [@greenwood_tree_2017]. However, in other cases these traits have failed to link to drought tolerance ([Maréchaux et al. 2019](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/1365-2745.13321)), and the direction of response is not always consistent; for instance, higher $WD$ has been associated with greater drought resistance at a global scale [@greenwood_tree_2017], but it correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011]. Thus, their perceived influence on drought resistance may actually reflect indirect correlations with other traits that more directly drive drought responses [@hoffmann_hydraulic_2011]. Recent work has shown a great potential for hydraulic traits to predict growth and mortality responses. Hydraulic traits including water potentials at which percent loss of conductivity surpass a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin have enabled prediction of drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics). More easily measurable leaf hydraulic traits with direct linkage to plant hydraulic function are emerging with potential to explain greater variation in plant distribution and function than the more commonly-measured $WD$ and $LMA$ [@medeiros_extensive_2019]. These include leaf area shrinkage upon desiccation ($PLA_{dry}$; [@scoffoni_leaf_2014]) and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016]. The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree performance under drought remains untested.

The long timeframe captured in tree-ring data enables us to address the question of whether tree size and species' traits have similar influence across different drought events, or whether that influence is more strongly predicted by community-level responses to variable drought severity, duration, and timing based on tree size and traits. Tree growth responses vary with drought characteristics such as the timing of onset, antecedent moisture conditions, drought duration, and temperature-driven evaporative demand [@dorangeville_drought_2018], but the question of how tree size and species' traits impact growth responses across droughts still remains. While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., @lloret_components_2011;  @dorangeville_drought_2018), they generally focus on species-level responses, and do not consider the roles of tree size and microenvironment. The ecological field-based studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on the most extreme droughts with dramatic impacts (*e.g.*, @allen_global_2010; @bennett_larger_2015; @stovall_tree_2019; @anderegg_meta-analysis_2016). Thus, our knowledge of forest responses to more modest but frequent droughts - e.g., those with historical return intervals of one to two decades - remains limited. There is evidence that the degree to which larger trees are more impacted by drought increases with the severity of drought conditions [@bennett_larger_2015; @stovall_tree_2019], so the influence of tree size may be weaker in less severe--but more common--droughts. Thus, while we expect many of the mechanisms shaping drought responses to be universal, we have little understanding of how tree size and traits interact with drought characteristics, and the extent to which their influence is consistent across droughts.

Here, to yield a functional understanding of how tree size, microenvironment characteristics, and species' traits collectively shape drought responses, we test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and forest census data from a large forest dynamics plot in Virginia (USA). First, we focus on the role of tree size and its interaction with microenvironment. We test whether, consistent with most forests globally, larger-diameter trees tend to have lower drought resistance ($Rt$) in this forest, which is in a region (eastern North America) represented by only two studies in @bennett_larger_2015. 
We then test hypotheses designed to disentangle the relative importance of tree height; crown exposure; and soil water availability, which should be greater for larger trees in dry but not in perpetually wet microsites. 
Second, we focus on the role of species' functional and hydraulic traits, testing the hypothesis that species’ traits-–particularly leaf hydraulic traits-–predict $Rt$. We test predictions that drought resistance is correlated with wood density--either postively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011]-- and positively correlated with specific leaf area, but that hydraulic leaf traits including $PLA_{dry}$ and $\pi_{tlp}$ are better predictors. We expect that ring-porous species would have greater drought resistance than semi-ring and diffuse-porous species, as has been previously observed [@kannenberg_linking_2019; @elliott_forest_2015; @friedrichs_species-specific_2009].

@@

Resistance was negatively correlated with $ln[TWI]$ (Tables 4-5), negating the idea that trees in moist microsites would be less affected by drought. Nevertheless, we tested for a negative $ln[H] *ln[TWI]$ interaction, which could indicate that smaller trees (with smaller rooting volume) are more susceptible to drought in drier microenvironments with a deeper water table. This hypothesis was rejected; the $ln[H] *ln[TWI]$ interaction was never significant and had a consistently positive coefficient (Table 4).

@@

Tree size, microenvironment, and hydraulic traits shaped tree growth responses across three droughts (Table 1). The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height rather than crown exposure [@liu_effect_1993; @stovall_tree_2019]. We found only a marginal additional effect of crown exposure, with a tendency for lowest $Rt$ among the most exposed (dominant) and suppressed trees. There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ (consistent with @zuleta_drought-induced_2017; @stovall_tree_2019), and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments. The negative effect of height on $Rt$ held after accounting for species' traits.  Drought sensitivity was not consistently linked to species' $LMA$, $WD$, or xylem type (ring- vs. diffuse porous), but was negatively correlated with the leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts. This is a novel finding in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses. The direction of responses was mostly consistent across droughts, supporting the conclusion that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts (Tables 4-5), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. These findings advance our knowledge of the factors that make trees vulnerable to growth declines during drought--and, by extension, likely make them more vulnerable to mortality (**Sapes et al. 2019, doi: 10.1093/treephys/tpz062**).

The droughts considered here were of a magnitude that has occurred with an average frequency of approximately once every 10-15 years (Fig. 1a, @helcoski_growing_2019) and had modest impacts on tree growth (Fig. 1b). These droughts were classified as severe (1977) or extreme (1966, 1999) according to the PDSI metric and have been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019]; however, extreme, multiannual droughts ("megadroughts") of the type that have triggered massive tree die-off in other regions (e.g., @allen_global_2010; @stovall_tree_2019) have not occurred in the Eastern United States within the past several decades [@clark_impacts_2016]. Of the droughts considered here, the 1966 drought, which was preceded by two years of dry conditions (Fig. S2), severely stressed a larger portion of trees (Fig. 1b). It may be notable that the tendency for large trees to have lowest resistance was most pronounced in this drought, consistent with other findings that this tendency increases with drought severity [@bennett_larger_2015; @stovall_tree_2019]. Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig. 1b), potentially due to decreased leaf area of competitors during the drought. It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-26% for each individual drought (Table 5).

Our analysis indicates that tree height has a stronger influence on drought response than does canopy position (Tables 1,4,5). This is consistent with, and reinforces, previous findings that biophysical constraints make it impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019]. However, this result must be interpreted with some caution, given that collinearity between the two variables (Fig. 2d) makes it impossible to confidently partition causality. Taller trees are more likely to be in dominant canopy positions (Fig. 2d) and, largely as a consequence of their position relative to others, face different microenvironments (Fig. 2a-b). Even under non-drought conditions, evaporative demand and maximum leaf temperatures increase with tree height [@smith_temperature_1977; @bretfeld_plant_2018; @kunert_revised_2017], and such conditions would incur extra stress during drought, when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves. However, some decoupling between height and canopy position is introduced by the configuration of neighboring trees (Fig. 2d) [@muller-landau_testing_2006], and height was an overall stronger predictor of drought response than crown position (Tables 1,4,5).

>>>>

**Authors:** 
Ian R. McGregor^1,2^,
Ryan Helcoski^1^,
Norbert Kunert^1,3^,
Alan J. Tepley^1,4^,
Erika B. Gonzalez-Akre^1^,
Valentine Herrmann^1^,
Joseph Zailaa^1,5^,
Atticus E.L. Stovall^1,6,7^,
Norman A. Bourg^1^,
William J. McShea^1^,
Neil Pederson^8^,
Lawren Sack^9,10^,
Kristina J. Anderson-Teixeira^1,3^*

@@

Forests play a critical global role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change [@friedlingstein_climatecarbon_2006]. An important aspect of this uncertainty lies with physiological responses of trees to drought [@kennedy_implementing_2019]. In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014]; [Dai et al. 2018](https://link.springer.com/article/10.1007/s40641-018-0101-6)), often despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015;  @cook_unprecedented_2015]. Droughts, intensified by climate change, have been affecting forests worldwide [@allen_global_2010], and are expected to continue as one of the most important drivers of forest change in the future [@allen_underestimation_2015]. Understanding and modeling forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought resistance, and the extent to which their influence is consistent across droughts. However, it has proved difficult to resolve the many factors affecting tree growth during drought with available forest census data, which only rarely captures extreme drought, and with tree-ring records, which rarely represent many species and size classes at a single site and rarely sampled in association with detailed ecological data. The aim of this study was to pair tree-ring and forest plot data to test how tree size and traits shaped growth responses to historical droughts.

@@

Debates have also arisen regarding the traits influencing tree growth responses to drought. Commonly-measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011](**Martin-Benito & Pederson 2015, doi:10.1111/jbi.12462**) and other forest biomes around the world [@greenwood_tree_2017]. However, in other cases these traits could not explain drought tolerance drought tolerance ([Maréchaux et al. 2019](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/1365-2745.13321)), or the direction of response was not always consistent; for instance, higher $WD$ has been associated with greater drought resistance at a global scale [@greenwood_tree_2017], but it correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011]. Thus, their perceived influence on drought resistance may actually reflect indirect correlations with other traits that more directly drive drought responses [@hoffmann_hydraulic_2011]. Recent work has shown a great potential for hydraulic traits to predict growth and mortality responses. Hydraulic traits including water potentials at which percent loss of conductivity surpass a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin have enabled prediction of drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics). More easily measurable leaf hydraulic traits with direct linkage to plant hydraulic function are emerging with potential to explain greater variation in plant distribution and function than the more commonly-measured $WD$ and $LMA$ [@medeiros_extensive_2019]. These include leaf area shrinkage upon desiccation ($PLA_{dry}$; [@scoffoni_leaf_2014]) and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016]. The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree performance under drought remains untested.

The long timeframe captured in tree-ring data enables us to address the question of whether tree size and species' traits have similar influence across different drought events, or whether that influence is more strongly predicted by community-level responses to variable drought severity, duration, and timing based on tree size and traits. Tree growth responses vary with drought characteristics such as the timing of onset, antecedent moisture conditions, drought duration, and temperature-driven evaporative demand [@dorangeville_drought_2018], but the question of how tree size and species' traits impact growth responses across droughts still remains. While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., @lloret_components_2011;  @dorangeville_drought_2018)([**Truettner et al**](https://doi.org/10.1016/j.foreco.2018.01.044)), they generally focus on species-level responses, and do not consider the roles of tree size and microenvironment. The ecological field-based studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on the most extreme droughts with dramatic impacts (*e.g.*, @allen_global_2010; @bennett_larger_2015; @stovall_tree_2019; @anderegg_meta-analysis_2016). Thus, our knowledge of forest responses to more moderate but frequent droughts - e.g., those with historical return intervals of one to two decades - remains limited. There is evidence that the degree to which larger trees are more impacted by drought increases with the severity of drought conditions [@bennett_larger_2015; @stovall_tree_2019], so the influence of tree size may be weaker in less severe--but more common--droughts. Thus, while we expect many of the mechanisms shaping drought responses to be universal, we have little understanding of how tree size and traits interact with drought characteristics, and the extent to which their influence is consistent across droughts.

Here, to yield a functional understanding of how tree size, microenvironment characteristics, and species' traits collectively shape drought responses, we test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three drought (1966, 1977, 1999), species functional and hydraulic trait measurements, and forest census data from a large forest dynamics plot in Virginia (USA). First, we focus on the role of tree size and its interaction with microenvironment. We test whether, consistent with most forests globally, larger-diameter trees tend to have lower drought resistance ($Rt$) in this forest, which is in a region (eastern North America) represented by only two studies in @bennett_larger_2015. 
We then test hypotheses designed to disentangle the relative importance of tree height; crown exposure; and soil water availability, which should be greater for larger trees in dry but not in perpetually wet microsites. 
Second, we focus on the role of species' functional and hydraulic traits, testing the hypothesis that species’ traits-–particularly leaf hydraulic traits-–predict $Rt$. We test predictions that drought resistance is correlated with wood density--either postively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011]-- and positively correlated with specific leaf area, but that hydraulic leaf traits including $PLA_{dry}$ and $\pi_{tlp}$ are better predictors. We expect that ring-porous species would have greater drought resistance than semi-ring and diffuse-porous species, as has been previously observed [@kannenberg_linking_2019; @elliott_forest_2015; @friedrichs_species-specific_2009].

@@

Resistance was negatively correlated with $ln[TWI]$ (Tables 4-5), rejecting the idea that trees in moist microsites would be less affected by drought. Nevertheless, we tested for a negative $ln[H] *ln[TWI]$ interaction, which could indicate that smaller trees (with smaller rooting volume) are more susceptible to drought in drier microenvironments with a deeper water table. This hypothesis was rejected as the $ln[H] *ln[TWI]$ interaction was never significant and had a consistently positive coefficient (Table 4).

@@

Tree size, microenvironment, and hydraulic traits shaped tree growth responses across three droughts at our study site (Table 1). The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height rather than crown exposure [@liu_effect_1993; @stovall_tree_2019]. We found only a marginal additional effect of crown exposure, with a tendency for lowest $Rt$ among the most exposed (dominant) and suppressed trees. There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ (consistent with @zuleta_drought-induced_2017; @stovall_tree_2019), and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments. The negative effect of height on $Rt$ held after accounting for species' traits.  Drought sensitivity was not consistently linked to species' $LMA$, $WD$, or xylem type (ring- vs. diffuse porous), but was negatively correlated with the leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts. This is a novel finding in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses. The direction of responses was mostly consistent across droughts, supporting the premise that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts (Tables 4-5), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. These findings advance our knowledge of the factors that make trees vulnerable to growth declines during drought--and, by extension, likely make them more vulnerable to mortality (**Sapes et al. 2019, doi: 10.1093/treephys/tpz062**).

The droughts considered here were of a magnitude that has occurred with an average frequency of approximately once every 10-15 years (Fig. 1a, @helcoski_growing_2019) and had modest impacts on tree growth (Fig. 1b). These droughts were classified as severe (1977) or extreme (1966, 1999) according to the PDSI metric and have been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019]; however, extreme, multiannual droughts or so-called “megadroughts” of the type that have triggered massive tree die-off in other regions (e.g., @allen_global_2010; @stovall_tree_2019) have not occurred in the Eastern United States within the past several decades [@clark_impacts_2016]. Of the droughts considered here, the 1966 drought, which was preceded by two years of dry conditions (Fig. S2), severely stressed a larger portion of trees (Fig. 1b). It may be notable that the tendency for large trees to have lowest resistance was most pronounced in this drought, consistent with other findings that this physiological response increases with drought severity [@bennett_larger_2015; @stovall_tree_2019]. Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig. 1b), potentially due to decreased leaf area of competitors during the drought. It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-26% for each individual drought (Table 5).

Our analysis indicates that tree height has a stronger influence on drought response than does canopy position (Tables 1,4,5). This is consistent with, and reinforces, previous findings that biophysical constraints make it impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019]. However, this result must be interpreted with some caution, given that collinearity between the two variables (Fig. 2d) makes it impossible to confidently partition causality. Taller trees are more likely to be in dominant canopy positions (Fig. 2d) and, largely as a consequence of their position relative to others, face different microenvironments (Fig. 2a-b). Even under non-drought conditions, evaporative demand and maximum leaf temperatures increase with tree height [@smith_temperature_1977; @bretfeld_plant_2018; @kunert_revised_2017], and such conditions would incur additional stress during drought, when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves. However, some decoupling between height and canopy position is introduced by the configuration of neighboring trees (Fig. 2d) [@muller-landau_testing_2006], and height was an overall stronger predictor of drought response than crown position (Tables 1,4,5). Belowground, taller trees would tend to have larger root systems, but the potentially greater access to water did not override the vulnerability conferred by height--and, in fact, greater moisture access in non-droought years (here, higher TWI) appears to make trees more vulnerable to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].
